---
title: "Dataset generation: 'vanilla' 1-CMT model"
author: "Niklas Hartung"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Scope

First, we look at the simplest possible model: 

* one compartment, a single i.v. bolus dose (= 1 ODE);
* identical sampling times across individuals;
* no covariates.

### Preliminaries

Required packages

```{r}
library(deSolve)
library(dplyr)
library(ggplot2)
```

Set a seed for reproducible results:

```{r}
set.seed(1)
```

### Model definition

Population and sample size

```{r}
N <- 1000                      # nb. of individuals (IDs)
t_obs <- c(0.5,1,2,4,8,16,24)  # sampling times
n_obs <- length(t_obs)         # nb. of samples per ID
```

Dosing scheme: i.v. bolus, single dose, flat

```{r}
D     <- 100                   # [mg] 
route <- "iv_bolus"
```

Hyperparameters: 

* fixed effects `fe`
* standard deviation of random effects `sd_re`
* residual unexplained variability standard deviation `sd_ruv` 

```{r}
fe     <- c(V  = 20,      # [L]
            CL =  3)      # [L/h]
sd_re  <- c(V  = 0.4,     # unitless
            CL = 0.3)     # unitless
sd_ruv <- 0.2             # [mg/L]
```

Individual parameters are log-normally distributed to ensure positivity:

```{r}
theta <- matrix(NA, nrow = N, ncol = 2, dimnames = list(NULL, c("V","CL")))
for (i in 1:N) {
    theta[i,"V"]  <- fe[["V"]]  * rlnorm(n = 1, meanlog = 0 , sdlog = sd_re[["V"]])
    theta[i,"CL"] <- fe[["CL"]] * rlnorm(n = 1, meanlog = 0 , sdlog = sd_re[["CL"]])
}
head(theta)
```

ODE right-hand side

```{r}
rhs <- function(t, y, param) {
    dX <- - param["CL"] / param["V"] * y
    list(dX)
}
```

Solving ODE and computing model prediction:

```{r}
y_pred <- vector(mode = "list", length = N)
for (i in 1:N) {
    theta_i <- theta[i, ]
    x0 <- c(pla=D)        # i.v. bolus: initial amount in plasma = dose
    
    # ODE solution: amount; observable: concentration (= amount/volume)
    x_sol <- ode(y = x0, 
                 times = c(0,t_obs),          # add dosing time as initial condition
                 func = rhs, 
                 parms = theta_i)[-1,"pla"]   # remove dosing time from ode solution
    y_pred[[i]] <- x_sol / theta_i["V"]
}
```

### Create, visualize and write data

The (hierarchical) data:

```{r}
df <- data.frame(
        ID    = rep(1:N, each = length(t_obs)),
        time  = rep(t_obs, N),
        y_obs = unlist(y_pred) + rnorm(n = N*n_obs, mean = 0, sd = sd_ruv),
        dose  = NA,
        route = NA
    ) |>
    rbind(data.frame(
        ID = 1:N,
        time = rep(0, N),
        y_obs = NA,
        dose = D,
        route = route
    )) |>
    arrange(ID,time)
```

Peek at the data:

```{r}
head(df, n = 20)
```

Visualization (just a subset of individuals, for easier visualization)

```{r}
ggplot(subset(df, ID <= 10 & !is.na(y_obs)), 
       aes(x=time,y=y_obs,group = ID)) + 
    geom_line() + 
    geom_point() +
    xlab("Time [h]") + 
    ylab("Concentration [mg/L]")
```

Write to file:

```{r}
write.csv(df, file = "../data/vanilla-1cmt.csv")
```

