---
title: "Dataset generation: 1-CMT model with loading/maintenance doses"
author: "Niklas Hartung"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Scope

We consider the following setting: 

* one compartment, a single i.v. infusion as loading dose, 
  followed by oral maintenance doses (= 3 ODEs);
* identical sampling times across individuals;
* no covariates.

### Preliminaries

Required packages

```{r}
library(deSolve)
library(logitnorm)
library(dplyr)
library(ggplot2)
```

Set a seed for reproducible results:

```{r}
set.seed(1)
```

### Model definition

Number of individuals (IDs) 

```{r}
N <- 1000
```

Dosing scheme: 1h i.v. infusion followed by oral doses

```{r}
infusion_dur <- 1    # [h]
infusion_amt <- 100
infusion_rate<- infusion_amt/infusion_dur # for use in ODEs
n_maintain   <- 4
maintain_amt <- 200
D            <- rep(c(infusion_amt, maintain_amt), 
                    times = c(1,n_maintain))    # [mg] 
tD           <- seq(0, by = 24, length = n_maintain + 1)    # in [h], daily dose
route        <- rep(c("iv_infusion","oral"), times = c(1,n_maintain)) 
```

Sampling times

```{r}
t_obs <- c(0.5,1,2,4,8, 24+c(2,8), 48+c(2,8), 72+c(2,8), 96+c(2,8))  # sampling times
n_obs <- length(t_obs)         # nb. of samples per ID
```

Hyperparameters: 

* fixed effects `fe`
* standard deviation of random effects `sd_re`
* residual unexplained variability standard deviation `sd_ruv` 

```{r}
fe     <- c(ka   = 0.2,   # [1/h]
            Fbio = 0.4,   # unitless
            V    = 20,    # [L]
            CL   =  3)    # [L/h]
sd_re  <- c(ka   = 0.1,   # unitless
            Fbio = 0.5,   # unitless
            V  = 0.4,     # unitless
            CL = 0.3)     # unitless
sd_ruv <- 0.2             # [mg/L]
```

Individual parameters are log-normally distributed to ensure positivity,
except for `Fbio`, which is logit-normally distributed such that it stays in [0,1]:

```{r}
theta <- matrix(NA, nrow = N, ncol = 4, dimnames = list(NULL, c("ka","Fbio","V","CL")))
for (i in 1:N) {
    
    theta[i,"ka"]   <- rlnorm(n = 1, meanlog = log(fe[["ka"]]),     sdlog = sd_re[["ka"]])
    theta[i,"Fbio"] <- rlogitnorm(n = 1,  mu = logit(fe[["Fbio"]]), sigma = sd_re[["Fbio"]])
    theta[i,"V"]    <- rlnorm(n = 1, meanlog = log(fe[["V"]]),      sdlog = sd_re[["V"]])
    theta[i,"CL"]   <- rlnorm(n = 1, meanlog = log(fe[["CL"]]),     sdlog = sd_re[["CL"]])
    
}
head(theta)
```

Forcing function encoding the infusion

```{r}
infusion_rate_function <- approxfun(x = c(0, infusion_dur), 
                                    y = c(infusion_rate, 0), 
                                    method = "constant", 
                                    rule = 2)
```

ODE right-hand side

```{r}
rhs <- function(t, y, param) {
    
    # Infusion only happens if there's drug in the bag
    infusion <- if (y[["bag"]] > 0) infusion_rate else 0
  
    dX <- c(
        -param[["ka"]] * y[["gut"]],                                  # gut 
         -infusion,                                                   # bag
         param[["ka"]] * param[["Fbio"]] * y[["gut"]] +               # plasma 
            infusion  - param[["CL"]] / param[["V"]] * y[["pla"]]     # plasma continued
    )
    list(dX)
}
```

Encoding repeated dosing via "events"

```{r}
eventdat <- data.frame(
    var  = unname(c("iv_infusion" = "bag", "oral" = "gut")[route]),
    time = tD,
    value = D,
    method = "add"
) |> 
    rbind(data.frame(
        var = "bag",
        time = infusion_dur,
        value = 0,
        method = "rep"
    )) |>
    arrange(time)
```
        
Solving ODE and computing model prediction:

```{r}
y_pred <- vector(mode = "list", length = N)
for (i in 1:N) {
    theta_i <- theta[i, ]
    x0 <- c(gut = 0, bag = 0, pla = 0)
    
    # event times need to be included in ODE solver times
    t_ode <- sort(c(tD,t_obs))

    # ODE solution: amount; observable: concentration (= amount/volume)
    x_sol <- ode(y = x0, 
                 times = t_ode,         
                 func = rhs,
                 parms = theta_i,
                 events = list(data = eventdat) # handling of repeated dosing
             )[t_ode %in% t_obs,"pla"]          # remove dosing times from ode solution
    y_pred[[i]] <- x_sol / theta_i["V"]
}
```

### Create, visualize and write data

The (hierarchical) data:

```{r}
df <- data.frame(
        ID    = rep(1:N, each = length(t_obs)),
        time  = rep(t_obs, N),
        y_obs = unlist(y_pred) + rnorm(n = N*n_obs, mean = 0, sd = sd_ruv),
        dose  = NA,
        route = NA,
        duration = NA
    ) |>
    rbind(data.frame(
        ID   = rep(1:N, each = n_maintain + 1),
        time = rep(tD, times = N),
        y_obs = NA,
        dose = rep(D, times = N),
        route = rep(route, times = N),
        duration = rep(ifelse(route == "iv_infusion", infusion_dur, NA), times = N)
    )) |>
    arrange(ID,time)
```

Peek at the data:

```{r}
head(df, n = 30)
```

Visualization (just a subset of individuals, for easier visualization),
including dosing cycles

```{r}
ggplot(subset(df, ID <= 10 & !is.na(y_obs)), 
       aes(x=time,y=y_obs,group = ID)) + 
    geom_line() + 
    geom_point() +
    geom_vline(xintercept = tD, color = "blue") +
    xlab("Time [h]") + 
    ylab("Concentration [mg/L]")    

```

Write to file:

```{r}
write.csv(df, file = "../data/loadingMaintainance.csv")
```

