---
title: "Dataset generation: 1-CMT model with oral dosing (Hamberg Warfarin PK model)"
author: "Niklas Hartung"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Scope 

We consider a (somewhat simplified) model for pharmacokinetics of Warfarin 
by Hamberg et al.: 

* one compartment with a single oral dose (=2 ODEs);
* different sampling times and numbers of samples across individuals;
* four continuous covariates (body weight and three variables relating to CYP2C9
  genotype, already converted from categorical to continuous).

### Preliminaries

Required packages

```{r}
library(deSolve)
library(dplyr)
library(ggplot2)
```

Set a seed for reproducible results:

```{r}
set.seed(1)
```

### Model definition

Population and sample size

```{r}
N <- 50                          # nb. of individuals (IDs)
tfull    <- c(8, 16, 24, 48, 72) # planned sampling times (in [h])
pMissing <- 0.1                  # probability that an observation is missing
nfull    <- length(tfull)        # number of planned samples
```

Individual sampling times (some samples may be missing and actual sampling times
may differ from the nominal ones)

```{r}
t_obs <- vector(mode = "list", length = N)
for (i in 1:N) {
    t_i <- tfull[as.logical(rbinom(n = nfull, size = 1, p = 1-pMissing))]  # missing samples
    t_i <- t_i + rnorm(n = length(t_i))                                    # actual sampling time
    t_obs[[i]] <- t_i
}
# look at the first few IDs:
head(t_obs)
```

Covariate distribution:

* draw body weight from a normal distribution (for simplicity)
* draw CYP2C9 alleles according to reported frequencies

```{r}
BWref <- 80
BWsd  <- 15
BW <- rnorm(n = N, mean = BWref, sd = BWsd)

#                     *1     *2    *3
pCYP2C9Alleles <- c(81.4,  11.2,  7.4) / 100
CYP2C9 <- rmultinom(n = N, size = 2, prob = pCYP2C9Alleles)
```

Dosing scheme: single oral flat dosing

```{r}
D     <- 3                   # [mg] 
route <- "oral"
```

Hyperparameters: 

* fixed effects `fe`
* standard deviation of random effects `sd_re`
* residual unexplained variability standard deviation `sd_ruv` 

```{r}
fe <- c(ka  = 0.2,          # [1/h]
        Vd  = 14.3,         # [L]
        CL1 =  0.174,       # [L/h]
        CL2 =  0.0879,      # [L/h]
        CL3 =  0.0422)      # [L/h]
sd_re <- c(Vd = 0.054,      # unitless
           CL = 0.089)      # unitless
sd_ruv <- 0.099             # [mg/L], error on log-scale
```

Individual parameters are affected by covariates:

* body weight on `Vd`;
* body weight and CYP2C9 genotype on `CL`

In addition, a log-normally distributed random effect (to ensure positivity) is applied.

```{r}
theta <- matrix(NA, nrow = N, ncol = 3, dimnames = list(NULL, c("Vd","CL","ka")))
for (i in 1:N) {
    relbw <- BW[i]/70

    theta[i,1] <- fe[["Vd"]] * relbw * rlnorm(n = 1, meanlog = 0, sdlog = sd_re[["Vd"]])
    theta[i,2] <- fe[c("CL1","CL2","CL3")] %*%  CYP2C9[,i] * relbw^0.75 * rlnorm(n = 1, meanlog = 0, sdlog = sd_re[["CL"]])
    theta[i,3] <- fe[["ka"]]
}
head(theta)
```

ODE right-hand side (assuming full bioavailability)

```{r}
rhs <- function(t, y, param) {
    dX <- c(
        -param[["ka"]] * y[["gut"]],
         param[["ka"]] * y[["gut"]] - param[["CL"]] / param[["Vd"]] * y[["pla"]]
    )
    list(dX)
}
```

Solving ODE and computing model prediction:

```{r}
y_pred <- vector(mode = "list", length = N)
for (i in 1:N) {
    theta_i <- theta[i, ]
    x0 <- c(gut = D, pla = 0)
    
    # ODE solution: amount in CMT 2; observable: concentration in CMT 2(= amount/volume)
    x_sol <- ode(y = x0, 
                 times = c(0,t_obs[[i]]),     # add dosing time as initial condition
                 func = rhs,  
                 parms = theta_i)[-1,"pla"]   # remove dosing time from ode solution
    y_pred[[i]] <- x_sol / theta_i[["Vd"]]
}
```


The (hierarchical) data:

```{r}
nobs <- vapply(t_obs,length,1)
df <- data.frame(
    ID    = rep(1:N, times = nobs),
    time  = unlist(t_obs),
    y_obs = unlist(y_pred) * rlnorm(n = sum(nobs), meanlog = 0, sdlog = sd_ruv),
    dose  = NA,
    route = NA
) |>
    rbind(data.frame(
        ID = 1:N,
        time = rep(0, N),
        y_obs = NA,
        dose = D,
        route = route
    )) |>
    arrange(ID,time)
```

Peek at the data:

```{r}
head(df, n = 20)
```

Visualization (just a subset of individuals, for easier visualization)

```{r}
ggplot(subset(df, ID <= 10), 
       aes(x=time,y=y_obs,group = ID)) + 
    geom_line() + 
    geom_point() +
    xlab("Time [h]") + 
    ylab("Concentration [mg/L]")
```

Write to file:

```{r}
write.csv(df, file = "../data/hamberg-warfarin-pk.csv")
```




### Dataset: 2-CMT model with nonlinear clearance

Now, we consider a more complex scenario:

* two-compartment model
* multiple dosing
* 

Study design

```{r}
N <- 1000        # nb. of individuals
nmax <- 10
tnom <- c()      # nominal observation times


sample(c(TRUE,FALSE), size = length(tnom), replace = TRUE, prob = c(0.8,0.2))

n <- rbinom(N, size = 10, prob = 0.8)

```





